---
title: "Forecasting the 2020 American Presidential Election"
author: "Ran Li, Andrei Velasevic"
date: "Monday November 2, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE, warning=FALSE}
library("tidyverse")
library("lme4")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# Loading in the cleaned survey Data
survey <- read_csv("vote.csv")

#Removing blank rows
survey_reduced = survey %>% drop_na()

## Add age group column
age_group <- ifelse(
  survey_reduced$age<=29 , "18-29",
  ifelse(
    survey_reduced$age<=44, "30-44",
    ifelse(
      survey_reduced$age<=64, "45-64",
      ifelse(
        survey_reduced$age>=65, "65+", NA
      )
    )
  )
)

survey_reduced = cbind(survey_reduced, age_group)
levels(factor(survey_reduced$age_group))


## Remove some data so that the vote 2020 category is binary
survey_reduced =
  survey_reduced %>% 
  filter(vote_2020 != "I am not sure/don't know") %>% 
  filter(vote_2020 != "I would not vote") %>% 
  filter(vote_2020 != "Someone else")


## Grouping race into a new category
race_group = ifelse(survey_reduced$race_ethnicity == "White",
                    "White",
                    ifelse(survey_reduced$race_ethnicity == "Black, or African American", "Black, or African American",
                           ifelse(survey_reduced$race_ethnicity =="Asian (Chinese)", "Asian (Chinese)",
                                  ifelse(survey_reduced$race_ethnicity =="Asian (Japanese)", "Asian (Japanese)",
                                         ifelse(survey_reduced$race_ethnicity =="Asian (Asian Indian)"|survey_reduced$race_ethnicity =="Asian (Filipino)"|survey_reduced$race_ethnicity =="Asian (Korean)"|survey_reduced$race_ethnicity =="Asian (Other)"|survey_reduced$race_ethnicity =="Asian (Vietnamese)"|survey_reduced$race_ethnicity =="Asian (Other)"|survey_reduced$race_ethnicity =="Pacific Islander (Guamanian)"|survey_reduced$race_ethnicity =="Pacific Islander (Native Hawaiian)"|survey_reduced$race_ethnicity =="Pacific Islander (Other)"|survey_reduced$race_ethnicity =="Pacific Islander (Samoan)", "other asian or pacific islander",
                                                ifelse(survey_reduced$race_ethnicity =="American Indian or Alaska Native", "American Indian or Alaska Native",
                                                       ifelse(survey_reduced$race_ethnicity =="Some other race", "Some other race", NA
                                                          )))))))
survey_reduced = cbind(survey_reduced, race_group)


## Grouping Employment into its own category
employment_status = ifelse(survey_reduced$employment=="Full-time employed"|survey_reduced$employment=="Part-time employed"|
survey_reduced$employment=="Self-employed"|survey_reduced$employment=="Homemaker",
"employed",
ifelse(survey_reduced$employment=="Unemployed or temporarily on layoff",
       "unemployed",
       ifelse(survey_reduced$employment=="Permanently disabled"|
                survey_reduced$employment=="Student"|
                survey_reduced$employment=="Retired", "not in labor force", "n/a")))

survey_reduced = cbind(survey_reduced, employment_status)
```

# Model
In this analysis, our goal is to predict the popular vote outcome of the 2020 American federal election (citation). To do this we are employing a post-stratification technique with a multilevel logistic regression model. The specifics of the model and post-stratification calculation will be discussed in the following subsection:

## Model Specifics
We decided to use multilevel logistic regression to model the proportion of voters who will vote for Donald Trump. In our model, we are using a variety of factors to model the probability of Donald Trump being voted for. The data for the model was taken from the Democracy Fund + UCLA Nationscape ‘Full Data Set’, which is an individual-level survey of *6479* observations of *18* categories. To create a meaningful model for our prediction, we took the explanatory variables to be:  
1. Gender\newline
2. Age\newline
3. State\newline
4. Education\newline
5. Race\newline
6. Employment\newline

and build a two_level logistic regression model given by:

$$
\begin{aligned}
P(Y_{ij}\in \{Donald\ Trump, Joe\ Biden\})= logit^{-1}(\beta_{0j} +\beta_1(gender)+\beta_2(age\ group)+\\\beta_3(employment)+\beta_4(education)+\epsilon_i)
\end{aligned}
$$

where $Y$ represents the proportion of voters who will vote for Donald Trump. $\beta_0$ represents the base intercept of the model and $\beta_1$ correspond to the slope of the model in relation to the gender category, treated as a binary outcome (Male or Female).So, in example for every possible outcome of gender, we expect a $\beta_1$ change in the probability of voting for Donald Trump.However, since we believe that people in similar age, state, education, race, employment group behave similar in voting, the intercept term is dependent on these varaibles and randomness of the intercept term is modeled by the second level regression model as followed:
$$\beta_{0j} =  r_{00} + r_{01}b_{j}^{state}+
r_{02}b_{j}^{race}+\epsilon$$
Where in the above model, $\beta_{0j}$, corresponds to the random intercept, $r_{00}$ is an overall intercept for second level regression and $r_{0i}$s are the slopes in relation to the other dependent observations $b_j$.
The reason how we set categories in these varaibles will be explained in more detail in the poststratification section.
```{r, echo = FALSE}

## Building a logistic regression model
model = glmer(vote_trump ~ factor(gender)+factor(age_group)+ (1|state)+(1|education)+(1|race_group)+factor(employment_status),data = survey_reduced, family = binomial)

```

## Post-Stratification 
In order to estimate the proportion of voters who will vote for Donald Trump our group needed to perform a post-stratification analysis. The basic idea is to split data into cells and take weighted average of estimates within each cell This technique is used usually to account for underrepresented groups in the population. (citation: https://www.stata.com/manuals13/svypoststratification.pdf) and this is very useful because it decreases non response bias and bias from underrepresented groups. It also decreases variance estimates of models.\newline
Here cells are created based off of different ages greater than 18 (4 categories), gender(2 categories), states (51 categories, including Disctrict of Columbia), education (10 cataegories), race(7 categories), and employment (4 categories) with all possible combinations which generate a total of 114,240 cells. 
The ages are chosen due to certain age groups being succeptable to certain political views. Young adults tend to go to universities where they can express themselves and possibly amplify the views of the institution. Gender can also be a factor in seeing how President Trump's views on certain gender-related issues differ from Biden's. States are an obvious choice since many states have certain affiliations with political parties. Education is important because individuals may have better understandings of more complicated political topics if they have a higher education. As we have seen so far, there have been many big events that have caught the media's attention in months leading up to the election that relate to racial differences. Due to this we decided to also include census data of racial groups. Employment is a large topic in politics as unemployment rates are often brought up.\newline
After splitting cells, by using the model described in the "Methods" subsection, we are estimating the proportion of voters in each cell and weight each proportion estimate (within each bin) by the respective population size of that bin, formula for calculation the overall estimate of proportition voting Donald Trump is presented here:\newline
$$\hat{y}^{ps}=\frac{\sum{N_j\hat{y_j}}}{\sum{N_j}}$$

## Result
```{r, echo =FALSE}
## Results
library(kableExtra)
result = as.data.frame(coef(summary(model)))
result %>%
  kbl(digits = 3, caption = "Coefficients based on multilevel regression model") %>%
  kable_material(c("striped", "hover"))
```

```{r, include=TRUE}

# Here I will perform the post-stratification calculation
library("datasets")
census = read.csv("census_data.csv")
names(census)[2] <- "gender"
names(census)[3] <- "state"
names(census)[4] <- "education"
names(census)[6] <- "employment_status"

#ratio = census$n/sum(census$n)
#census = cbind(census, ratio)
census$gender = str_to_title(census$gender)

prediction <- model %>% predict(census, allow.new.levels = TRUE, type = "response")
census = cbind(census, prediction)
y_s = sum(census$prediction*census$n)/sum(census$n)
y_s
```

According to our calculations, we estimated the proportion of voters to be `r y_s`. Meaning approximately `r 100*round(y_s, 4)`% of people will vote for Donald Trump. By means of post-stratification, we calculated this number as the proportion of voters in favor of Donald J. Trump, modeled by a multilevel logistic regression. Our model which accounted for gender, age, race, education, and employment status gives us a clear answer in what we are searching for.

## Descrptive Analysis about Prediction Reuslt
We are also interested in predicting how different states behave in voting. By setting the population as each state, we can use the same formula mentioned before to  do prediction
```{r}
census
state_prediction = census %>% group_by(state) %>% summarize(Prediction = sum(n * prediction)/sum(n))
gender_prediction = census %>% group_by(gender) %>% summarize(Prediction = sum(n * prediction)/sum(n))
race_prediction = census %>% group_by(race_group) %>% summarize(Prediction = sum(n * prediction)/sum(n))
education_prediction = census %>% group_by(education) %>% summarize(Prediction = sum(n * prediction)/sum(n))
employment_prediction = census %>% group_by(employment_status)%>% summarize(Prediction = sum(n * prediction)/sum(n))
age_prediction = census %>% group_by(age_group) %>% summarize(Prediction = sum(n * prediction)/sum(n))

ggplot(state_prediction, aes(x=state, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")
```

```{r}

ggplot(age_prediction, aes(x=age_group, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")

ggplot(race_prediction,aes(x=race_group, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")

ggplot(gender_prediction,aes(x=gender, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")

ggplot(education_prediction,aes(x=education, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")

ggplot(employment_prediction, aes(x=employment_status, y=Prediction, group = 1)) + geom_line() + geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")

```




# Discussion


## Weaknesses


## Next Steps


# References

tidyverse:

  Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software,
  4(43), 1686, https://doi.org/10.21105/joss.01686
  
  
knitr:

  Yihui Xie (2020). knitr: A General-Purpose Package for Dynamic Report Generation in
  R. R package version 1.29.

  Yihui Xie (2015) Dynamic Documents with R and knitr. 2nd edition. Chapman and
  Hall/CRC. ISBN 978-1498716963

  Yihui Xie (2014) knitr: A Comprehensive Tool for Reproducible Research in R. In
  Victoria Stodden, Friedrich Leisch and Roger D. Peng, editors, Implementing
  Reproducible Computational Research. Chapman and Hall/CRC. ISBN
  978-1466561595


lme4:

  Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear
  Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48.
  doi:10.18637/jss.v067.i01.
  

datasets:

  R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.
  
  
Sample Data:

Press, Courier &amp;, et al. “New: Second Nationscape Data Set Release.” Democracy Fund Voter Study Group, Democracy Fund Voter Study Group, 30 Oct. 2020, www.voterstudygroup.org/publication/nationscape-data-set. 
  
  
Census Data:

Team, MPC UX/UI. “U.S. CENSUS DATA FOR SOCIAL, ECONOMIC, AND HEALTH RESEARCH.” IPUMS USA, University of Minnesota, Minnesota Population Center, usa.ipums.org/usa/index.shtml. 




“Poststratification —Poststratification for Survey Data.” Www.stata.com, 1996–2020 StataCorp LLC, www.stata.com/manuals13/svypoststratification.pdf. 
  

## Post-Stratification 
In order to estimate the proportion of voters who will vote for Donald Trump our group needed to perform a post-stratification analysis. The basic idea is to split data into cells and take weighted average of estimates within each cell This technique is used usually to account for underrepresented groups in the population. (citation: https://www.stata.com/manuals13/svypoststratification.pdf) and this is very useful because it decreases non response bias and bias from underrepresented groups. It also decreases variance estimates of models.\newline
Here cells are created based off of different ages greater than 18 (4 categories), gender(2 categories), states (51 categories, including Disctrict of Columbia), education (10 cataegories), race(7 categories), and employment (4 categories) with all possible combinations which generate a total of 114,240 cells. 
The ages are chosen due to certain age groups being succeptable to certain political views. Young adults tend to go to universities where they can express themselves and possibly amplify the views of the institution. Gender can also be a factor in seeing how President Trump's views on certain gender-related issues differ from Biden's. States are an obvious choice since many states have certain affiliations with political parties. Education is important because individuals may have better understandings of more complicated political topics if they have a higher education. As we have seen so far, there have been many big events that have caught the media's attention in months leading up to the election that relate to racial differences. Due to this we decided to also include census data of racial groups. Employment is a large topic in politics as unemployment rates are often brought up.\newline
After splitting cells, by using the model described in the "Methods" subsection, we are estimating the proportion of voters in each cell and weight each proportion estimate (within each bin) by the respective population size of that bin, formula for calculation the overall estimate of proportition voting Donald Trump is presented here:\newline
$$\hat{y}^{ps}=\frac{\sum{N_j\hat{y_j}}}{\sum{N_j}}$$

## Result
```{r, echo =FALSE}
## Results
library(kableExtra)
result = as.data.frame(coef(summary(model)))
result %>%
  kbl(digits = 3, caption = "Coefficients based on multilevel regression model") %>%
  kable_material(c("striped", "hover"))


## Age group and Vote
table_age = table(factor(survey_reduced$age_group), survey_reduced$vote_2020)
prop_age = as.data.frame(prop.table(table_age, 1))
names(prop_age)[1]=('Age')
names(prop_age)[2]=('Vote')
names(prop_age)[3]=('Ratio')

ggplot(data=prop_age, aes(x=Age, y=Ratio, fill=Vote)) + geom_bar(stat="identity")
```

```{r, include=TRUE}

# Here I will perform the post-stratification calculation
library("datasets")
census = read.csv("/Users/ranli/Desktop/STA304 PS3/census_data.csv")
names(census)[2] <- "gender"
names(census)[3] <- "state"
names(census)[4] <- "education"
names(census)[6] <- "employment_status"

#ratio = census$n/sum(census$n)
#census = cbind(census, ratio)
census$gender = str_to_title(census$gender)

model$
prediction <- model %>% predict(census, allow.new.levels = TRUE, type = "response")
census = cbind(census, prediction)
y_s = sum(census$prediction*census$n)/sum(census$n)
y_s
```

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).


# References




Where $y$ represents the proportion of voters who will vote for Donald Trump. Similarly, $\beta_0$ represents the intercept of the model, and is the probability of voting for Donald Trump at age 0. Additionally, $\beta_1$ represents the slope of the model. So, for everyone one unit increase in age, we expect a $\beta_1$ increase in the probability of voting for Donald Trump.



# Results

## Descriptive Statistics


According to our calculations, we estimated the proportion of voters to be `r y_s`. Meaning approximately `r 100*round(y_s, 4)`% of people will vote for Donald Trump. By means of post-stratification, we calculated this number as the proportion of voters in favor of Donald J. Trump, modeled by a multilevel logistic regression. Our model which accounted for gender, age, race, education, and employment status gives us a clear answer in what we are searching for.

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).


# References
